---
title: "Class 9 Halloween Mini Project"
author: "Katelyn (PID: A16682595)"
format: pdf
---

# Importing Candy Data

```{r}
candy_data <- "candy-data.csv"
candy <- read.csv(candy_data, row.names = 1)
head(candy)
```

*Q1. How many different candy types are in this dataset?*
```{r}
nrow(candy)
```


*Q2. How many fruit candy types are in the dataset?*
```{r}
sum(candy$fruity)
```



# What is Your Favorite Candy?

*Q3. What is your favorite candy in the dataset and what is it’s winpercent value?*
```{r}
candy["Whoppers",]$winpercent
```


*Q4. What is the winpercent value for “Kit Kat”?*
```{r}
candy["Kit Kat",]$winpercent
```


*Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?*
```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```



```{r}
# Loading and using the skimr package
library("skimr")
skim(candy)
```


*Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?*

Win percent goes all the way up to 82.18, but all other columns are binary(either 0 or 1) or between 0 and 1.


*Q7. What do you think a zero and one represent for the candy$chocolate column?*

0 = "no", 1 = "yes"


*Q8. Plot a histogram of winpercent values*
```{r}
library(ggplot2)
ggplot(candy, aes(winpercent)) +
  geom_histogram(binwidth = 10)
```


*Q9. Is the distribution of winpercent values symmetrical?*

Not quite. It looks a little skewed.


*Q10. Is the center of the distribution above or below 50%?*

The center of the distribution looks below 50%.


*Q11. On average is chocolate candy higher or lower ranked than fruit candy?*
```{r}
# Turning chocolate into T/F values
choc.Inds <- as.logical(candy$chocolate)
# Identifying the win percents of chocolate candy
choc.win <- candy[choc.Inds,]$winpercent
# Taking the average win percent of chocolate
mean(choc.win)

# Doing the same for fruit candy
fruit.Inds <- as.logical(candy$fruity)
fruit.win <- candy[fruit.Inds,]$winpercent
mean(fruit.win)

#Is chocolate candy higher ranked than fruit candy on average?
mean(choc.win) > mean(fruit.win)
```

*Q12. Is this difference statistically significant?*
```{r}
t.test(choc.win, fruit.win)
```
Yes. The p-value is really small, indicating the difference is significant.



# Overall Candy Rankings

*Q13. What are the five least liked candy types in this set?*
```{r}
library(dplyr)
candy %>%
  arrange(winpercent) %>%
  head(5)
```

*Q14. What are the top 5 all time favorite candy types out of this set?*
```{r}
candy %>%
  arrange(desc(winpercent)) %>%
  head(5)
```

*Q15. Make a first barplot of candy ranking based on `winpercent` values.*
```{r}
ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```

*Q16. Use the `reorder()` function to get the bars sorted by `winpercent`*
```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col()
```

```{r}
# Adding a color vector, mycols
mycols <- rep("black", nrow(candy))
mycols[as.logical(candy$chocolate)] <- "chocolate"
mycols[as.logical(candy$bar)] <- "brown"
mycols[as.logical(candy$fruity)] <- "pink"
```

```{r}
# Generating a colored graph
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col(fill = mycols)
```

*Q17. What is the worst ranked chocolate candy?*

Sixlets

*Q18. What is the best ranked fruity candy?*

Starburst

# Taking a Look at Pricepercent

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=mycols) + 
  geom_text_repel(col=mycols, size=3.3, max.overlaps = 5)
```

*Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?*

Reese's miniatures are the most bang for your buck.

*Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?*
```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```
Nik L Nips are the least popular and one of the most expensive.

```{r}
# Pricepercent ggplot:
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_point() +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), xend = 0), col = "grey40")
```

# Exploring the Correlation Structure

```{r}
library(corrplot)
#calculate correlation
cij <- cor(candy)
#plot correlation
corrplot(cij)
```

*Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?*

Chocolate and fruity are strongly anti-correlated.

*Q23. Similarly, what two variables are most positively correlated?*

Chocolate being a bar and having a high win percent are most positively correlated.

# Principal Component Analysis

```{r}
pca <- prcomp(candy, scale = T)
summary(pca)
```

```{r}
# Comparing PC1 with PC2 in basic R:
plot(pca$x[,1:2], col = mycols, pch = 16)
```

```{r}
# Make a new dataframe for ggplot to read
my_data <- cbind(candy, pca$x[,1:3])

# PC2 vs. PC1 in ggplot:
p <- ggplot(my_data, aes(PC1, PC2, size = winpercent/100, text=rownames(my_data),
            label=rownames(my_data))) +
  geom_point(col = mycols)
p
```

```{r}
# Adding text labels with ggrepel:
p + geom_text_repel(size=3.3, col=mycols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
# Looking at our PCA loadings:
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

*Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?*

Fruity, hard and pluribus. This makes sense because fruity candy tend to be hard and come in multiples. 
