---
title: "Week 12: RNAseq Galaxy"
author: "Katelyn Wei (PID: A16682595)"
format: pdf
---

# Part 4: Population Scale Analysis

~230 gene samples were processed from a population and placed in the file "PopAnalysis.txt". 

Q13: Determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

Let's first look at what we've got.
```{r}
x <- read.table("PopAnalysis.txt")
```

There are 462 samples in this table based on the `nrow()` function:
```{r}
nrow(x)
```

We can use the `sum()` function and logicals to find the sample size of each genotype:
```{r}
# Sample size A/A
sum(x[,2] == "A/A")

# Sample size A/G
sum(x[,2] == "A/G")

# Sample size G/G
sum(x[,2] == "G/G")
```

Now how do we get the median of each? Using the dplyr package, we can `group_by()` genotype then use `summarize()` to find median expression levels:
```{r}
library(dplyr)

x %>%
  group_by(geno) %>%
  summarize(medianGeno = median(exp))
```



> Q14: Generate a boxplot with a box per genotype. What could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP affect the expression of ORMDL3?

```{r}
# Boxplot using ggplot
library(ggplot2)

ggplot(x, aes(geno, exp, fill = geno)) +
  geom_boxplot() +
  geom_point(alpha = 0.5) +
  labs(x = "Genotype", y = "Expression Level") +
  theme(legend.position = "none")
```

Based on this plot, you can infer that A/A tend to have higher expression levels than G/G. The SNP seems to affect ORMDL3 expression.