---
title: "Class 5 Data Visualization with ggplot2"
author: "Katelyn Wei (PID: A16682595)"
format: pdf
---

## Using GGPLOT

The ggplot2 package does not come installed with R. Use the `install.package()` function to do so.

```{r}
head(cars)
```

To use ggplot I need to load up before I can call any of the functions in the package. Use the `library()` function to do this. 

```{r}
library(ggplot2)
ggplot()
```

All ggplot figures have at least 3 things: 
- data(the stuff we want to plot)
- aesthetic mapping(assigned visual aspects to your data)
- geoms (how our dataset will be visualized)

```{r}
ggplot(cars) + 
  aes(x=speed, y=dist) +
  geom_point()
```

ggplot is not the only graphing system in R - there are lots of others. There are even "base R" graphics (remember the plot function?).

```{r}
plot(cars)
```

## Lab 5

### 6. Creating Scatterplots

#### Cars Plot

Adding a trendline: 

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() +
  geom_smooth()
```

straightening the line and removing the shaded region:

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

Adding labels and removing color:

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title="Car Braking Distances", x="Speed (mph)", y="Distance (ft)", caption="Dataset: 'cars'") +
  theme_bw()
```

#### Genes Plot

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

Basic Scatterplot: 
```{r}
p <- ggplot(genes) +
  aes(x = Condition1, y = Condition2) +
  geom_point()
p
```

Adding/changing the colors and adding labels:

```{r}
p + aes(color=State) +
  scale_color_manual(values = c("blue", "gray", "red")) + 
  labs(title = "Gene Expression Changes Upon Drug Treatment", x = "Control (no drug)", y = "Drug Treatment")
```

### 7. Going Further

```{r}
library(gapminder)
library(dplyr)
```

Basic Scatterplot:

```{r}
gapminder_2007 <- gapminder %>% filter(year==2007)
ggplot(gapminder_2007) +
  aes(x=gdpPercap, y=lifeExp) +
  geom_point()
```

1957 Data Plot:

```{r}
gapminder_1957 <- gapminder %>% filter(year==1957)
ggplot(gapminder_1957) +
  aes(x=gdpPercap, y=lifeExp, color=continent, size=pop) +
  geom_point(alpha=0.7) +
  scale_size_area(max_size = 15)
```

Comparing 1957 to 2007:

```{r}
gapminder_1957 <- gapminder %>% filter(year==1957 | year==2007)
ggplot(gapminder_1957) +
  aes(x=gdpPercap, y=lifeExp, color=continent, size=pop) +
  geom_point(alpha=0.7) +
  scale_size_area(max_size = 15) +
  facet_wrap(~year)
```

